---
title: "lab_08"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key

```{r echo=FALSE}
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2019 5-year estimates so you can identify the codes you need:

**Task** load the ACS variables and save them to a dataframe

```{r}
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2019 5-year estimates? You will need to find the variable for median age first.
**A1**

```{r}
```

**Q2** Which Maryland jurisdictions have a median age of at least 50?
**A2**

```{r}
```

**Q3** Considering the answers to the first two questions, what kinds of public needs do places with the demographic characteristics of Baltimore City have compared to the counties from Q2? What kinds of programs or services should the latter jurisdictions expect to spend more money on?
**A3**

**Q4** Which two Maryland jurisdictions have the most children between 5 and 17 that speak Spanish at home? Can you tell which one has the most?
**A4**

```{r}
```

**Q5** A better answer to Q4 involves calculating percentages of children between 5 and 17 that speak Spanish at home. Which counties have the highest percentage of 5-17 year-olds that speak Spanish at home? Start with the code you wrote for Q4 and adapt it so that you also load the total population of 5-17 year-olds by county into another dataframe. In each case, rename the `estimate` columns to the name you've given the ACS codes, remove the `variable` columns in each using select and then join them based on GEOID and NAME. Then you can add a column to calculate the percentage of the total 5-17 population that speaks Spanish at home.
**A5**

```{r}
```

**Q6** Given the counties with the largest percentage of 5-17 year-olds that speak Spanish at home, what stands out to you? How would you describe the top 5 counties? How would you describe Garrett County's numbers?
**A6**


**Q7** Which Maryland jurisdiction has the largest percentage of its households without internet access? Hint: the total households variable is `DP02_0001`. For this question, you'll need to load households data and the no internet data into separate dataframes, rename the `variable` columns in each and remove the variable columns in each using select and then join them based on GEOID and NAME. Then you can add a column to calculate the percentage of households without internet.
**A7**

```{r}
```

**Q8** Of all the Maryland jurisdictions listed in your answer to Q5, which one is the outlier? Write a couple of sentences describing why you chose that jurisdiction and how you might frame a story pitch based on this data.
**A8** 
