---
title: "lab_08"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API Key

You'll need your Census API Key:

```{r echo=FALSE}
census_api_key("YOUR KEY HERE", install=TRUE)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the 2019 5-year estimates? You will need to find the variable for median age first.
**A1** Baltimore City, with an estimated median age of 35.4 years and a margin of error of 0.2 years.

```{r}
md_median_age <- get_acs(geography='county', variables=c(median_age='B01002_001'), state='MD')
arrange(md_median_age, estimate)
```

**Q2** Which Maryland jurisdictions have a median age of at least 50?
**A2** Talbot and Worcester counties

```{r}
md_median_age %>% filter(estimate >= 50)
```

**Q3** Considering the answers to the first two questions, what kinds of public needs do places with the demographic characteristics of Baltimore City have compared to those of Talbot and Worcester counties? What kinds of programs or services should those jurisdictions expect to spend more money on?
**A3** 

**Q4** Which two Maryland jurisdictions have the most children between 5 and 17 that speak Spanish at home?
**A4** Prince George's and Montgomery counties, although given the margin of error it's not clear which one has the most.

```{r}
speak_spanish_5_17 <- get_acs(geography='county', variables=c(speak_spanish_5_17='B16007_004'), state='MD')
arrange(speak_spanish_5_17, desc(estimate))
```

**Q5**
**A5**
