---
title: "lab_04_sample"
author: "sean mussenden"
date: "8/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

This lab is accumulative. It is designed to test methods and concepts you've learned in prior weeks, with special emphasis on things learned in this week's prelab chapters.    

## The frame

You are a data journalist at a national news organization. New U.S. Census figures were released today showing demographic change in the U.S. between 2010 and 2020.  Your editor has asked you to produce a data notebook with the most interesting growth trends you can find, which can be used as the foundation for a fully-reported story. You have the following hypotheses, based on interviews, 
## How to complete this lab

To complete this lab, you will need to do one or more of the following task:
* Run prewritten code in an existing codeblock.
* Modify prewritten code in an existing codeblock and run.
* Write code in a preexisting codeblock and run.
* Create a new codeblock, write code and run. 
* Write out an answer to a question in the form of a complete sentence.
* Develop a question to answer.  

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')

library(tidyverse)
```

## Load Data

We're going to load the slice of PPP loans for West Virginia.

**Task** Run the codeblock below to load the data.

```{r}

national_race_decennial_census_demographics_2010_2020 <- read_rds("national_race_decennial_census_demographics_2010_2020.rds")


decennial_census_demographics_2010_2020 <- read_rds("decennial_census_demographics_2010_2020.rds")


```


## Explore for now

What happened to the share of the population for each demographic group between 2010 and 2020?

```{r}



x <- national_race_decennial_census_demographics_2010_2020 %>%
  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
  mutate(total_pop_2010 = sum(race_pop_2010),
         total_pop_2020 = sum(race_pop_2020)) %>%
  mutate(share_of_total_2010 = race_pop_2010/total_pop_2010*100) %>%
  mutate(share_of_total_2020 = race_pop_2020/total_pop_2020*100) %>%
  mutate(percentage_share_change_2010_2020 = share_of_total_2020-share_of_total_2010)

```

Which state saw the largest percentage point decline in the share of the population that is white. 

```{r}
y <- x <- national_race_decennial_census_demographics_2010_2020 %>%
  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
  mutate(total_pop_2010 = sum(race_pop_2010),
         total_pop_2020 = sum(race_pop_2020)) %>%
  mutate(share_of_total_2010 = race_pop_2010/total_pop_2010*100) %>%
  mutate(share_of_total_2020 = race_pop_2020/total_pop_2020*100) %>%
  mutate(percentage_share_change_2010_2020 = share_of_total_2020-share_of_total_2010)

```


What states had biggest pop growth. Which had smallest. 
https://www.washingtonpost.com/nation/interactive/2021/census-population-changes/
https://www.nytimes.com/interactive/2021/08/12/us/2020-census-race-ethnicity.html
## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.
**Q1** How many rows and columns does this dataframe have?
**A1** [write your answer here]

```{r}
glimpse(wv_ppp_loans)
```

## Answer questions

**Q2** Which West Virginia county got the most loans? How many?
**A2** [write your answer here]

```{r}
# Write the code to produce your answer here

```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** [write your answer here]

```{r}
# Write the code to produce your answer here

```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** [write your answer here]


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q6** What West Virginia politician (with his family) owns the company identified in question 2? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** [write your answer here]

**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**A7** [write your answer here]
```{r}
# Write the code to produce your answer here
```

**Q8** Produce a summary table that has the following information for each West Virginia county:
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_county_summary` so that you can examine it with the data viewer.

What newsworthy nuggets or patterns jump out at you that you could imagine dropping into a news story or doing follow up reporting on? Write up a few sentence that summarizes one county.

**A8** [write your answer here]
```{r}
# Write the code to produce your answer here
```
