---
title: "lab_04_sample"
author: "sean mussenden"
date: "8/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

This lab is accumulative. It is designed to test methods and concepts you've learned in prior weeks, with special emphasis on things learned in this week's prelab chapters.    

## The frame

You are a data journalist at a national news organization. New U.S. Census figures were released today showing demographic change in the U.S. between 2010 and 2020.  Your editor has asked you to produce a data notebook with the most interesting growth trends you can find, which can be used as the foundation for a fully-reported story. You have the following hypotheses, based on interviews, 
## How to complete this lab

To complete this lab, you will need to do one or more of the following **Tasks**:
* Run prewritten code in an existing codeblock.
* Modify prewritten code in an existing codeblock and run.
* Write code in a preexisting codeblock and run.
* Create a new codeblock, write code and run. 
* Write out an answer to a question in the form of a complete sentence.
* Develop a question to answer.  

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')

library(tidyverse)
```

## Load Data

We're going to load the slice o

```{r}
national_race_decennial_census_demographics_2010_2020 <- read_rds("national_race_decennial_census_demographics_2010_2020.rds")


decennial_census_demographics_2010_2020 <- read_rds("decennial_census_demographics_2010_2020.rds")


```

## Answer questions

**Q1** Which racial category had the largest percent increase between 2010 and 2020? Modify the code below by completing the empty `mutate()` and `arrange()` functions to produce a table with the answer.
**A1** The "two or more races" category grew by 275 percent, a nearly 4X increase.  
# https://www.census.gov/library/stories/2021/08/improved-race-ethnicity-measures-reveal-united-states-population-much-more-multiracial.html

```{r}

pct_change_by_category <- national_race_decennial_census_demographics_2010_2020 %>%
  mutate() %>%
  arrange()

pct_change_by_category

### Key
#pct_change_by_category <- national_race_decennial_census_demographics_2010_2020 %>%
#  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
#  arrange(desc(pct_change_2010_2020))

```

**Q2** Which racial category saw a decline in population between 2010 and 2020? Write code below to produce a table with the answer.
**A2** The white population declined by nine percentage points between 2010 and 2020. 
```{r}

### Key
#pct_change_by_category <- national_race_decennial_census_demographics_2010_2020 %>%
#  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
#  arrange(pct_change_2010_2020)


```

**Q3** Write a lede for a news story based on the tables produced in questions 1 and 2. Your lede should be one or two sentences. You may find it helpful to scan this page before writing.  https://www.census.gov/library/stories/2021/08/improved-race-ethnicity-measures-reveal-united-states-population-much-more-multiracial.html
**A3** The number of Americans who identify as multiracial jumped sharply between 2010 and 2020, while the "white alone" population declined, an analysis of new U.S. Census data shows. Adjustments to Census questionnaires were partly responsible for the changes, the agency said. [NOTE: we'll have them read this story this week https://www.npr.org/2021/08/22/1029609786/2020-census-data-results-white-population-shrinking-decline-non-hispanic-race]

**Q4** Which Maryland county had the biggest raw number increase in total population between 2010 and 2020? Which county had the biggest loss? 
**A4** Prince George's County had the biggest increase. Baltimore County had the biggest loss. 

```{r}
maryland_counties <- decennial_census_demographics_2010_2020 %>%
  filter(state == "Maryland") %>%
  mutate(change_total_pop_2010_2020 = (total_2020-total_2010))

```

**Q5** Which Maryland county had the biggest percentage increase in total population between 2010 and 2020? Which county had the biggest percentage loss? 
**Q5** Frederick County had the biggest increase. Allegany County had the largest loss.

```{r}
maryland_counties <- decennial_census_demographics_2010_2020 %>%
  filter(state == "Maryland") %>%
  mutate(pct_change_total_pop_2010_2020 = (total_2020-total_2010)/total_2010*100)

```

**Q6** What percentage of Maryland's population lived in Montgomery County in 2020?
**Q6** 17 percent.

```{r}
maryland_counties <- decennial_census_demographics_2010_2020 %>%
  filter(state == "Maryland") %>%
  mutate(maryland_total_2020 = sum(total_2020)) %>%
  mutate(share_of_maryland_pop_2020 = total_2020/maryland_total_2020*100)

```

**Q7** Using all of your powers, identify the most interesting lede that you can about changes in the Hispanic population in Maryland between 2010 and 2020.  You'll need to start by generating a few questions, write code to perationalize those questions and synthesis the results. 
**Q7** 
* Maryland's Hispanic population exploded over the last decade, growing at nearly 10 times the rate as the state as a whole, an analysis of new U.S. Census data showed. 
* Every single county in the state saw its Hispanic population increase, including the seven Maryland counties that saw their overall population shrink. 
* The share of the state's population that identifies as Hispanic grew from one-in-12 in 2010 to one-in-8 in 2020. 
* Maryland's Hispanic population is concentrated in Montgomery and Prince George's County.  Though only a third of Maryland's population lives in those two large suburban D.C. counties, together they house more than 60 percent of the state's Hispanic population.

```{r}

maryland_hispanic <- decennial_census_demographics_2010_2020 %>%
  select(county, state, contains("total"), hispanic_2010,hispanic_2020) %>%
  filter(state == "Maryland") %>%
  mutate(maryland_total_2020 = sum(total_2020),
         maryland_hispanic_2020 = sum(hispanic_2020),
         maryland_total_2010 = sum(total_2010),
         maryland_hispanic_2010 = sum(hispanic_2010)
         ) %>%
  mutate(overall_hispanic_percentage_2020 = maryland_hispanic_2020/maryland_total_2020,
         overall_hispanic_percentage_2010 = maryland_hispanic_2010/maryland_total_2010) %>%
  mutate(share_of_maryland_total_2020 = total_2020/maryland_total_2020*100,
         share_of_maryland_hispanic_2020 = hispanic_2020/maryland_hispanic_2020*100,
         share_of_maryland_total_2010 = total_2010/maryland_total_2010*100,
         share_of_maryland_hispanic_2010 = hispanic_2010/maryland_hispanic_2010*100,
         total_change_2010_2020 = (total_2020-total_2010)/total_2010*100,
         hispanic_change_2010_2020 = (hispanic_2020-hispanic_2010)/hispanic_2010*100,
         ) %>%
  mutate(pct_change_maryland_hispanic_2010_2020 = (maryland_hispanic_2020-maryland_hispanic_2010)/maryland_hispanic_2010*100,
         pct_change_maryland_total_2010_2020 = (maryland_total_2020-maryland_total_2010)/maryland_total_2010*100)

```

```{r}
**Q4** What might be the biggest loss. 
https://www.census.gov/library/stories/2021/08/improved-race-ethnicity-measures-reveal-united-states-population-much-more-multiracial.html

Filtering
Sorting
Mutating pct/change, difference, pct of total

## 


## Explore for now
# SAMPLE LAB 05 AGGREGATES

What happened to the share of the population for each demographic group between 2010 and 2020?

```{r}



x <- national_race_decennial_census_demographics_2010_2020 %>%
  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
  mutate(total_pop_2010 = sum(race_pop_2010),
         total_pop_2020 = sum(race_pop_2020)) %>%
  mutate(share_of_total_2010 = race_pop_2010/total_pop_2010*100) %>%
  mutate(share_of_total_2020 = race_pop_2020/total_pop_2020*100) %>%
  mutate(percentage_share_change_2010_2020 = share_of_total_2020-share_of_total_2010)

```

Which state saw the largest percentage point decline in the share of the population that is white. 

```{r}
y <- x <- national_race_decennial_census_demographics_2010_2020 %>%
  mutate(pct_change_2010_2020 = (race_pop_2020-race_pop_2010)/race_pop_2010*100) %>%
  mutate(total_pop_2010 = sum(race_pop_2010),
         total_pop_2020 = sum(race_pop_2020)) %>%
  mutate(share_of_total_2010 = race_pop_2010/total_pop_2010*100) %>%
  mutate(share_of_total_2020 = race_pop_2020/total_pop_2020*100) %>%
  mutate(percentage_share_change_2010_2020 = share_of_total_2020-share_of_total_2010)

```


Which county in Maryland saw the biggest


What states had biggest pop growth. Which had smallest. 
https://www.washingtonpost.com/nation/interactive/2021/census-population-changes/
https://www.nytimes.com/interactive/2021/08/12/us/2020-census-race-ethnicity.html
## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.
**Q1** How many rows and columns does this dataframe have?
**A1** [write your answer here]

```{r}
glimpse(wv_ppp_loans)
```

## Answer questions

**Q2** Which West Virginia county got the most loans? How many?
**A2** [write your answer here]

```{r}
# Write the code to produce your answer here

```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** [write your answer here]

```{r}
# Write the code to produce your answer here

```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** [write your answer here]


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q6** What West Virginia politician (with his family) owns the company identified in question 2? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** [write your answer here]

**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**A7** [write your answer here]
```{r}
# Write the code to produce your answer here
```

**Q8** Produce a summary table that has the following information for each West Virginia county:
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_county_summary` so that you can examine it with the data viewer.

What newsworthy nuggets or patterns jump out at you that you could imagine dropping into a news story or doing follow up reporting on? Write up a few sentence that summarizes one county.

**A8** [write your answer here]
```{r}
# Write the code to produce your answer here
```
