---
title: "lab_04"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Load Data

Let's keep working with Maryland loan applications contained in the PPP data.

**Task** Download the file from the above link and run the codeblock below to load the data.

```{r}
maryland_ppp <- read_csv('lab_04.rds')
```

To answer the questions below, you'll need to join both the NAICS code data and the ZCTA demographic data from the book chapter. There's a particular set of steps you'll need to perform to make that work properly.

First, let's load [the NAICS code data we have](https://umd.box.com/s/thou3merdzwzbdimbvd06zkax1zndcsk):

**Task** Download the file from the above link and run the codeblock below to load the data.

```{r}
naics_codes <- read_csv('data/naics_codes.csv')
```

Next join the two and then create a 5-character `zip5` column that we can use to join to the ZCTA data:

**Task** Run the codeblock below to join the data.

```{r}
maryland_ppp_with_naics <- maryland_ppp %>% left_join(naics_codes)
maryland_ppp_with_naics <- maryland_ppp_with_naics %>% mutate(zip5 = str_sub(zip, 1, 5))
```

Let's add the Zip Code Tabulation Area dataset. Remember that we'll need to make the `ZCTA5N` column a character column, not a numeric one:

**Task** Run the codeblock below to load the data.

```{r}
maryland_zcta <- read_csv('data/maryland_zcta.csv')
maryland_zcta <- maryland_zcta %>% mutate(across(ZCTA5N, as.character))
```

Now we can join the PPP data to the ZCTA data:

**Task** Run the codeblock below to join the data.

```{r}
maryland_ppp_with_naics_and_demographics <- maryland_ppp_with_naics %>% left_join(maryland_zcta, by=c("zip5"="ZCTA5N"))
```

## Answer questions

**Q1** What are the top industries (using the NAICS title) for PPP applications in Maryland?
**A1** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q2** What are the top industries (using the NAICS title) for PPP applications in Maryland with amounts of more than $150,000?
**A2** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q3** Which Maryland zip code with at least 20 percent non-Hispanic Asian population had the largest sum of PPP loans? Also show the number (count) of loans.
**A3** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q4** Which Maryland zip code had the largest sum of loans where the reported race of the applicant was "Asian", and what is the percentage of non-Hispanic Asian population in that zip code? Also show the number (count) of loans.
**A4** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q5** What might explain why 20850 has a significant Asian population (23.5%) and only 240 of the 3,186 loans (7.5%) are to Asian applicants?
**A5** [write your answer here]


**Q6** Are there Maryland zip codes that have loans with Asian applicants but do not appear to have Asian population in the ZCTA data? What's going on here?
**A6** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q7** How many industries have exactly one loan in the data (hint: you can filter _after_ using `summarise` to create a count). From a newsworthiness standpoint, which of these is most worthy of further exploration, and why?
**A7** [write your answer here]

```{r}
# Write the code to produce your answer here

```
